---
title: "Mammals and Water Availability: Plots Study Results"
description: |
    The aim of this study is to assess the effect of water scarcity on the distribution of an African herbivore guild along the Ruaha River, TZ
author:
    - name: "Stephanie Kramer-Schadt"
      url: https://ecodynizw.github.io  
      affiliation: Leibniz Institute for Zoo and Wildlife Research
      affiliation_url: https://izw-berlin.de/en/
      orcid_id: 0000-0002-9269-4446
    - name: "Cédric Scherer"
      url: https://cedricscherer.com
      affiliation: Leibniz Institute for Zoo and Wildlife Research
      affiliation_url: https://izw-berlin.de/en/
      orcid_id: 0000-0003-0465-2543
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  distill::distill_article:
    highlight: kate      
    code_folding: false   
    toc: true            
    toc_depth: 4         
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, 
  dev = "ragg_png", fig.width = 9, fig.height = 6, dpi = 600, retina = 1
)
Sys.setlocale("LC_TIME", "C")
```

* **Research question:** temporal change in group_size/ counts and distance to water
* **Study area:** Ruaha NP, TZ
* **Data:** Survey data by Claudia Stommel 2011-2013

---

# Setup

## Libraries

```{r libraries}
## statistical analyses
library(MuMIn) 
library(ggeffects)
library(jtools)
library(DHARMa)
library(glmmTMB)
library(mgcv)
library(broom.mixed)

## helpers
library(here)
library(dplyr)
library(tidyr)
library(readr)
library(tidymv)

## viz 
library(d6)
library(ggplot2)
library(RColorBrewer)
library(rcartocolor)
library(ggdensity)
library(prismatic)
library(patchwork)
library(magick)
```


## Theme

```{r theme}
# ggplot theme
theme_set(theme_d6())
```


## Data

```{r loaddata}
df_transects <- readRDS(here::here("output", "stommel_mastertable_20230127_cleaned.rds"))

head(df_transects)
names(df_transects)
#summary(df_transects)
```


## Define Labels

```{r set-labels}
# species sorting alphabetically according to their feeding type
# B        CD       E           G       I      K        Wb      Wh    Z
# grazer   omni     mixed       browser mixed  browser  grazer  omni  grazer

# Grazer:  B, Wb, Z
# mixed:   E, I,
# browser: G, K
# omni:    CD, Wh

species_labels <- as_labeller(
  c(`B`  = "Buffalo",
    `Wb` = "Waterbuck",
    `Z`  = "Zebra",
    `E`  = "Elephant",
    `I`  = "Impala",
    `G`  = "Giraffe",
    `K`  = "Kudu",
    `Wh` = "Warthog",
    `CD` = "Common duiker")
)

river_comp_labels <- as_labeller(
  c(`0`  = "upper",
    `1`  = "along",
    `2`  = "perpend")
)
```


## Define Color Codes

```{r set-colors}
## create table matching colors to species, grouped by feeding type
df_spec_colours <- data.frame(
  species = levels(df_transects$species), ## is orderd by feedtype (see above)
  color = c(
    brewer.pal(name = "YlGn", n = 9)[c(5,7,9)], ## grazer (3)
    carto_pal(name = "Burg", n = 7)[c(4,6)],    ## mixed (2)
    brewer.pal(name = "Blues", n = 7)[c(4,6)],  ## browser (2)
    carto_pal(name = "Purp", n = 7)[c(4,6)]     ## omni (2)
  )
)

my_spec_colours <- df_spec_colours$color
names(my_spec_colours) <- df_spec_colours$species


## create feeding type colors, as middle color of species colors above
my_feedtype_colours <- c(
  brewer.pal(name = "YlGn", n = 9)[7],  ## grazer
  brewer.pal(name = "Blues", n = 7)[5], ## browser
  carto_pal(name = "Burg", n = 7)[5],   ## mixed
  carto_pal(name = "Purp", n = 7)[5]    ## omni
)

## create color code for the 3 river classes (downstream/midstream/upstream)
my_river_class_cols <- c( "cadetblue1", "navyblue","lightblue") 
my_alongside_cols   <- c( "navyblue","cadetblue1") 

## create color code for seasons
my_season_cols      <- c("#637D37", "#E8CF55")

## create color code for distances
my_dist_class_cols  <- c("cadetblue1", "peachpuff4") 

## 
colors <- c("#FF0000", "#0000FF")
color_ramp <- colorRampPalette(colors)(6)
my_all_in_one_cols <- color_ramp
my_transect_cols <- viridis::turbo(n = 10)
```


# Main Tables

## Table 1 

Summarize the total counts per river section (perpendicular/ alongside). This might be misleading, as the same species were counted at every sampling event (33 weeks across the 3 years). 

```{r table-1}
data_table_1 <- 
  df_transects |> 
  group_by(transect_GRR, year, species, feed_type_A_char, dry_season) |> 
  summarize(
    sum_species_count = sum(species_count),
    mean_group_size   = round(mean(species_count), digits = 2),
    sd_group_size     = round(sd(species_count), digits = 2),
    mean_species_distance = round(mean(distance_to_water), digits = 2),
    sd_species_distance   = round(sd(distance_to_water), digits = 2)
  )

data_table_1

table_1 <- 
  data_table_1 |>
  select(
    transect_GRR ,year,species,feed_type_A_char, dry_season,
    sum_species_count, mean_group_size, mean_species_distance
  ) |>
  pivot_wider(
    names_from = c(year,dry_season),
    values_from = c(sum_species_count, mean_group_size, mean_species_distance)
  )

write.table(
  x = table_1,
  file = here("output", "stommel_transects_table_1.txt"),
  sep = '\t', col.names = TRUE, row.names = FALSE
)
```



# Main Figures


## Figure 1: Map Study Area

Separate .rmd file: ./R/01_map_fig1.rmd


## Figure 2: Herd Sizes

```{r fig-2, message=FALSE, fig.width=20*0.3937, fig.height=15*0.3937}
data_table_2 <- 
  df_transects |> 
  group_by(year, week, transect_GRR, dry_season, feed_type_A_char, species) |> 
  summarize(
    sum_species_count     = sum(species_count),
    mean_group_size       = round(mean(species_count), digits = 2),
    sd_group_size         = round(sd(species_count), digits = 2),
    mean_species_distance = round(mean(distance_to_water), digits = 2),
    sd_species_distance   = round(sd(distance_to_water), digits = 2)
  )

#summary(data_table_2)

ggplot(data_table_2, aes(x = transect_GRR, y = sum_species_count, fill = dry_season)) + 
  geom_boxplot(width = .7, color = "black", linewidth = .2, outlier.size = .3,
               position = position_dodge(preserve = "single")) + 
  facet_wrap(~ species, scale = "free_y", ncol = 3, 
             labeller = labeller(species = species_labels)) +
  scale_y_continuous(expand = expansion(mult = c(.005, .1)), limits = c(0, NA)) +
  scale_fill_manual(values = my_season_cols, name = "Dry season:") + 
  labs(x = "Transect section", 
       y = "Total count of individuals per sighting") +
  scale_x_discrete(labels = c("alongs.", "perpend.")) +
  theme_d6(grid = "y") +
  theme(#axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.text.x = element_text(angle = 35, vjust = 1, hjust = 1), 
        panel.spacing.x = unit(1, "lines"))

ggsave(here("plots", "Fig2_herd_sizes_transect_GRR.png"), 
       width = 20, height = 15, units = "cm", dpi = 600, bg = "white")
```


## Figure 3: Median Distances per Species and Transect GRR

```{r fig-3, fig.width=15*0.3937, fig.height=12*0.3937}
ggplot(df_transects, aes(x = dry_season, y = distance_to_water, fill = species)) + 
  geom_boxplot(
    aes(fill = species),
    width = 1, color = "grey90", outlier.shape = 21
  ) + 
  facet_wrap(~ transect_GRR, scale = "free_y", labeller = labeller(species = species_labels)) +
  scale_y_continuous(
    labels = scales::label_comma(), limits = c(0, NA),
    expand = expansion(mult = c(.001, .05))
  ) +  
  scale_fill_manual(
    values = clr_darken(my_spec_colours, .3), name = "Species:",
    guide = guide_legend(byrow = TRUE, nrow = 1, label.position = "bottom")
  ) +
  labs(x = "Dry season", y = "Distance to water [m]") +
  theme_d6(grid = "y") +
  theme(legend.key.width = unit(2, "lines"),
        legend.justification = "left")

ggsave(here("plots", "Fig3_distance_to_water_class.png"), 
       width = 15, height = 12, units = "cm", dpi = 600, bg = "white")
```



# Supplement Materials

## Table S2: Sightings per Species

Summarise number of sightings (not counts) per species per season for reviewer

```{r summary-sightings}
sight <- table(df_transects$dry_season, df_transects$species)
sight <- table(df_transects$dry_season, df_transects$species, df_transects$transect_GRR)
sight
```


## Figure S1: GAM

```{r, fig-s1-gam, fig.width=21*0.3937, fig.height=15*0.3937}
gam_mod <- gam(
  distance_to_water_rounded ~  species +  s(juldate, by = species, k = 3),
  family = nb(),
  data = df_transects
)

#summary(gam_mod)

#R-sq.(adj) =  0.113   Deviance explained = 10.4%
#-REML =  30895  Scale est. = 1         n = 3686

model_p <- predict_gam(gam_mod)

model_p |>
  ggplot(aes(juldate, fit, fill = species)) +
  geom_smooth_ci(species, linewidth = 1.2) +
  geom_line(aes(group = species), lwd = 2.5, color = "white") + 
  geom_smooth_ci(species, linewidth = 1.2) +
  labs(title = NULL, x = "Day since June 1st", y = "Smoothing function") +
  scale_color_manual(values = my_spec_colours, name = "Species:") +
  scale_fill_manual(values = my_spec_colours, name = "Species:") +
  scale_linetype_discrete(name = "Species:") +
  guides(color = guide_legend(nrow = 1))

ggsave(here("plots", "FigS1_model_fits_gam.png"), 
       width = 21, height = 15, units = "cm", dpi = 600, bg = "white") 
```


## Figure S2: Seasonal Densitiy Distributions

```{r prep-fig-densities}
## to calculate densities we need individual sighting data
df_transects_uncount <- tidyr::uncount(df_transects, species_count)

## function to create density plot as grid for all or selected species 
plot_densities <- function(spec = "all") {
  if (any(spec == "all")) { 
    dat <- df_transects_uncount
  } else {
    dat <- filter(df_transects_uncount, species %in% spec)
  }
  
  ggplot(dat, aes(lon, lat , fill = dry_season)) +
    geom_hdr() +
    geom_hdr_lines(aes(color = dry_season), lwd = .5) + 
    geom_point(shape = 21, size = .1) +
    geom_hdr_rug() +
    facet_wrap(~ species, labeller = labeller(species = species_labels)) +
    scale_color_manual(values = clr_darken(my_season_cols, .2), name = NULL) +
    scale_fill_manual(values = my_season_cols, name = NULL) +
    labs(x = "Longitude", y = "Latitude") +
    theme_d6(grid = "none")
}
```

All species for supplement figure

```{r, fig-s2-all, fig.width=20*0.3937, fig.height=20*0.39370}
plot_densities(spec = "all")

ggsave(here("plots", "FigS2_density_plots_all.png"), 
       width = 20, height = 20, units = "cm", dpi = 600, bg = "white")
```

Selected species for main figure: Zebra, Wb and B

```{r, fig-s2-selected, fig.width=20*0.3937, fig.height=9*0.3937}
plot_densities(spec = c("Z", "Wb", "B"))

# ggsave(here("plots", "FigS2_density_plots_selected.png"), 
#        width = 20, height = 9, units = "cm", dpi = 600, bg = "white")
```


## Figure S3: Mean Group Sizes 

```{r, fig-s3, fig.width=15*0.3937, fig.height=15*0.3937}
data_table_3 <- 
  df_transects |> 
  group_by(fyear, week, dry_season,species) |> 
  summarize(
    sum_species_count     = sum(species_count),
    mean_group_size       = round(mean(species_count), digits = 2),
    sd_group_size         = round(sd(species_count), digits = 2),
    mean_species_distance = round(mean(distance_to_water), digits = 2),
    sd_species_distance   = round(sd(distance_to_water), digits = 2)
  )

#summary(data_table_3)

ggplot(data_table_3, aes(x = dry_season, y = mean_group_size, fill = species)) + 
  geom_boxplot(width = .5, color = "black", position = position_dodge(width = 1)) +   
  facet_wrap(~ species, scale = "free_y", ncol = 3, 
             labeller = labeller(species = species_labels)) +
  scale_y_continuous(expand = expansion(mult = c(.005, .1)), limits = c(0, NA)) +
  scale_fill_manual(values = my_spec_colours, guide = "none") +
  labs(x = "Dry season", y = "Average group size per sampling week") +
  theme_d6(grid = "y")

ggsave(here("plots", "FigS3_mean_group_sizes.png"), 
       width = 20, height = 20, units = "cm", dpi = 600, bg = "white")
```



## Figure S4 + S5: Model Results

### GLMM model fit

-> models were run on outsourced file 5_b_DataAnalysis_sourcedModels_20230619.rmd

```{r}
## not run
# source(here("R", "5_b_DataAnalysis_sourcedModels_20230619.rmd"))
```

Load best model:

```{r model-best}
rds <- here("output", "mod444_20230619.rds")

## for now always run model due to version issues across computers
if (!file.exists(rds)) {
  mvvss <- glmmTMB(
    distance_to_water_rounded ~ species * (juldate + transect_GRR) + (1 | ftransect_id) ,
    data = df_transects, family = nbinom2, na.action = na.fail)
  
  write_rds(mvvss, file = rds)
} else {
  mvvss <- read_rds(rds) 
}
```

```{r model-results}
mod_summary <- summary(mvvss)
#mod_summary

mod_res   <- as.data.frame(round(mod_summary$coefficients$cond, digits = 7))
mod_names <- dimnames(mod_summary$coefficients$cond)[[1]]

res_table <- cbind(mod_names,mod_res)
#res_table

write.table(
  x = res_table,
  file = here("output", "model_feed_results_20230716.txt"),
  sep = '\t', col.names = TRUE, row.names = FALSE
)
```


### Figure S4: Model Diagnostics with DHARMa

```{r DHARMa}
testDispersion(mvvss)
simulationOutput <- simulateResiduals(fittedModel = mvvss, plot = FALSE)

png(here("plots", "FigS4b_dharma_residuals.png"), width = 700, height = 500)
plot(simulationOutput)
dev.off()
# 
# testOutliers(simulationOutput, type = "bootstrap")
# 
# png(here("plots", "FigS4a_dharma_prediction_cat.png"), width = 700, height = 400)
# testCategorical(simulationOutput, catPred = df_transects$species) 
# dev.off()
# 
# testCategorical(simulationOutput, catPred = df_transects$transect_GRR)
# 
# testZeroInflation(simulationOutput)
```


### Figure S5: Regression Coefficient Plots

```{r fig-s5, fig.width=20*0.3937, fig.height=23*0.3937}
jtools::plot_summs(mvvss, inner_ci_level = .75) +
  labs(y = NULL) +
  theme_d6(grid = "x", mono = "x") +
  theme(axis.text.y = element_text(hjust = 0),
        panel.border = element_blank(),
        axis.ticks = element_blank())

ggsave(here("plots", "FigS5_model_effects.png"), width = 20, height = 23, unit = "cm", bg = "white")

MuMIn::r.squaredGLMM(mvvss)
```


### Figure S6: GLMM Random Effects

```{r fig-s6, fig.width=15*0.3937, fig.height=12*0.3937}
re <- ranef(mvvss)

dd <- as.data.frame(re)

ggplot(dd, aes(y = grp, x = condval)) +
  geom_point() + 
  geom_errorbarh(aes(xmin = condval - 2 * condsd, xmax = condval + 2 * condsd), height = 0) +
  geom_vline(xintercept = 0, color = 'black', lty = 2) +
  scale_y_discrete(expand = c(0.05, 0)) +
  labs(x = "conditional effect", y = "transect ID",
       title = "Intercept") 


## alternative version
dd |> 
  mutate(
    lwr = condval - 2 * condsd,
    upr = condval + 2 * condsd,
    ci = if_else(lwr < 0 & upr > 0, TRUE, FALSE)
  ) |> 
  ggplot(aes(y = grp, x = condval)) +
  geom_point(aes(color = ci), size = 2) + 
  geom_errorbarh(aes(xmin = lwr, xmax = upr, color = ci), height = 0) +
  geom_vline(xintercept = 0, lty = 2) +
  scale_y_discrete(expand = c(0.05, 0)) +
  scale_color_manual(values = c("firebrick", "dodgerblue"), guide = "none") +
  labs(x = "conditional effect", y = "transect ID",
       title = "Intercept") 
   
ggsave(here("plots", "FigS6_random_effects_transID.png"), 
       width = 15, height = 12, units = "cm", dpi = 600, bg = "white")
```



# Figures for Reviewers

## Heatmap Sampling Transects

Summarise the time intervals at which each transect was sampled for reviewer.

```{r heatmap-time-intervals}
table(df_transects$weeks_all[df_transects$year == 2011], df_transects$ftransect_id[df_transects$year == 2011])

sy <- 2011 ## transect drive year
td <- table(df_transects$juldate[df_transects$year == sy], df_transects$ftransect_id[df_transects$year == sy])
m_td <- as.matrix(td)
colMain <- colorRampPalette(brewer.pal(8, "Blues"))(25)
heatmap(m_td, Colv = NA, Rowv = NA, scale = "column",col = colMain,
        xlab = "transect", ylab = "day since onset of dry season")
```


# Old Figures

## Silhoutte Plot Study Species

```{r  plot-phylopics, fig.width=15*0.3937, fig.height=2*0.3937}
## paths to images
zebra <- image_read("https://images.phylopic.org/images/81caf94e-5cbe-4e5e-8101-545abea2bfc0/raster/512x358.png")
kudu <- image_read("https://images.phylopic.org/images/e590147d-93c5-4d81-bbbf-bd1410cf8135/raster/1536x1055.png?build=159")
elephant <- image_read("https://images.phylopic.org/images/80db1004-bc9f-4318-84e7-cdd9639a1f3e/raster/1536x952.png?build=159")
giraffe <- image_read("https://images.phylopic.org/images/bbce74cf-4df3-4b7d-8b1d-f5b24dd3264a/raster/144x190.png")
duiker <- image_read(here("plots", "commonduiker_pathed_rohering.png"))
warthog <- image_read("https://images.phylopic.org/images/b1a641cd-25df-4dd0-865f-1a2728bd267e/raster/1536x916.png?build=159")
buffalo <- image_read("https://images.phylopic.org/images/65c4a9b3-dcde-4f0f-9a1f-8d71e74be9ec/raster/174x127.png?build=159")
impala <- image_read("https://images.phylopic.org/images/e07d1491-1d85-4c47-9f7d-075ea57bf0c5/raster/1536x1079.png?build=159")
waterbuck <- image_read("https://images.phylopic.org/images/f93103f1-e2a0-4c73-b274-c7b51afe4db0/raster/829x1024.png?build=159")

## import images
img <- function(png) { grid::rasterGrob(png, interpolate = FALSE) }

## create graphic
animals <- wrap_plots(
  img(zebra), img(waterbuck),img(buffalo),
  img(impala),img(elephant),
  img(giraffe),img(kudu),   
  img(warthog),img(duiker),   
  nrow = 1
)

animals + plot_layout(widths = c(0.8, 1, 1.4, 1.2, 1.4, 0.8, 1.3, 0.8, 0.8))

# ggsave(here("plots", "animals.png"), 
#        width = 15, height = 2, units = "cm", dpi = 600, bg = "white")
``` 


## Model Fits per Species

```{r model-fits-species, fig.width=30*0.3937, fig.height=15*0.3937}
mydfss1 <- ggpredict(mvvss, terms = c("juldate", "species",  "transect_GRR [alongside]")) 
mydfss2 <- ggpredict(mvvss, terms = c("juldate", "species", "transect_GRR [perpendicular]"))

plot_predict <- function(data, upper_y, title) {
  ggplot(data = data, aes(x = x, y = predicted)) + 
    geom_ribbon(
      aes(ymin = conf.low, ymax = conf.high, group = group), 
      fill = "white"
    ) + 
    geom_ribbon(
      aes(ymin = conf.low, ymax = conf.high, fill = group, 
          fill = after_scale(clr_desaturate(clr_lighten(fill, .2), .3)),
          color = after_scale(clr_lighten(fill, .6))), 
      alpha = .15, lwd = .2
    ) + 
    geom_line(aes(group = group), lwd = 2, color = "white") + 
    geom_line(aes(color = group), lwd = 1.2) + 
    coord_cartesian(expand = FALSE, ylim = c(0, upper_y)) +
    scale_y_continuous(labels = scales::label_comma(suffix = " km", scale = 1 / 1000)) +
    scale_color_manual(values = my_spec_colours, name = "Species:") +
    scale_fill_manual(values = my_spec_colours, name = "Species:") +
    guides(color = guide_legend(nrow = 1), fill = guide_legend(nrow = 1)) +
    labs(x = "Days since June 1st", y = "Distance to water", title = title) +
    theme(legend.key.width = unit(2, "lines"))
}

p1 <- plot_predict(mydfss1, upper_y =  3000, title = "Alongside transects")
p2 <- plot_predict(mydfss2, upper_y = 13000, title = "Perpendicular transects") +
        scale_y_continuous(labels = scales::label_comma(suffix = " km", scale = 1 / 1000), 
                           breaks = 0:4*3000) + 
        labs(y = NULL)

(p1 + p2) * theme(plot.title.position = "panel") + 
  plot_layout(guides = "collect")

# ggsave(here("plots", "Fig4_negbin_prediction_mod444_both.png"), 
#        width = 30, height = 15, units = "cm", dpi = 600, bg = "white")
```


# References

```{r}
citation("ggplot2")
citation("glmmTMB")
citation("DHARMa")
```


***

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
git2r::repository()
sessionInfo()
```

</details>
